---
title: "PBMC scRNA-seq Atlas Sensitivity Screen for COVID-19, Influenza, and Healthy Controls"
author: "James Hughes"
date: "2026-09-05"
output:
  html_document:
    df_print: paged
  pdf_document: default
params:
  data_dir: "data/raw"
header-includes:
  - \usepackage{float}
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  dev = "png",
  dpi = 150,
  cache = FALSE
)


options(scipen = 999)
RNGkind("L'Ecuyer-CMRG")
set.seed(1234)

repo_root <- normalizePath(if (file.exists("R/validation.R")) "." else "..", mustWork = TRUE)
knitr::opts_knit$set(root.dir = repo_root)
source(file.path(repo_root, "R/validation.R"))
source(file.path(repo_root, "R/runtime.R"))
input_data_dir <- normalizePath(params$data_dir, winslash = "/", mustWork = TRUE)
geo_metadata <- read.csv(file.path(repo_root, "data/metadata/geo_samples.csv"),
                         check.names = FALSE, stringsAsFactors = FALSE)
results_dir <- file.path(repo_root, "outputs/atlas_sensitivity/figures")
exports_dir <- file.path(repo_root, "outputs/atlas_sensitivity/tables")
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(exports_dir, recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(warning = TRUE, fig.path = paste0(results_dir, "/"))

save_plot <- function(plot_obj, filename, width = 7, height = 5, dpi = 200) {
  ggplot2::ggsave(
    filename = file.path(results_dir, filename),
    plot = plot_obj,
    width = width,
    height = height,
    dpi = dpi,
    bg = "white"
  )
}

fmt_int <- function(x) {
  format(as.integer(round(x)), big.mark = ",", scientific = FALSE, trim = TRUE)
}

fmt_num <- function(x, digits = 1) {
  format(round(x, digits), nsmall = digits, scientific = FALSE, trim = TRUE)
}
```

# Introduction

This unvalidated sensitivity workflow rebuilds only the global clustering and annotation stage of the PBMC scRNA-seq analysis. It is designed to produce a stable, defensible global atlas before any downstream composition testing, differential expression, pathway analysis, or subclustering.

# Methods

## Setup

```{r libraries, message=FALSE, warning=FALSE}
library(scDblFinder)
library(Seurat)
library(tidyverse)
library(Matrix)
library(patchwork)
library(SingleR)
library(SingleCellExperiment)
library(future)
library(celldex)
BiocParallel::register(BiocParallel::SerialParam())

available_workers <- parallel::detectCores(logical = FALSE)
available_workers <- if (is.na(available_workers)) 2 else available_workers
marker_workers <- max(1, min(2, available_workers - 1))
future::plan(sequential)
options(future.globals.maxSize = 8 * 1024^3)
```

## Load Data

```{r load-data}
counts <- Read10X(
  data.dir = input_data_dir
)

obj <- CreateSeuratObject(
  counts = counts,
  project = "COVID_PBMC_scRNA",
  min.cells = 3,
  min.features = 200
)

library_metadata <- map_library_metadata(colnames(obj), geo_metadata)
obj$sample <- library_metadata$sample
obj$group <- factor(library_metadata$group, levels = c("Normal", "Flu", "COVID"))
obj$geo_accession <- library_metadata$geo_accession
# These are library IDs. They must not be used as independent donor IDs.

genes_initial <- nrow(obj)
cells_initial <- ncol(obj)
```

## QC and Filtering

```{r qc}
obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT-")

qc_plot <- VlnPlot(
  obj,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  ncol = 3
)
save_plot(qc_plot, "atlas_qc_violin.png", width = 10, height = 4)
qc_plot

scatter1 <- FeatureScatter(obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
scatter2 <- FeatureScatter(obj, feature1 = "nCount_RNA", feature2 = "percent.mt")

save_plot(scatter1, "atlas_qc_scatter_counts_features.png", width = 5, height = 4)
save_plot(scatter2, "atlas_qc_scatter_counts_mito.png", width = 5, height = 4)

scatter1
scatter2
```

```{r filtering}
obj <- subset(
  obj,
  subset = nFeature_RNA > 250 &
    nFeature_RNA < 4500 &
    nCount_RNA < 25000 &
    percent.mt < 14
)

cells_after_qc <- ncol(obj)
genes_after_qc <- nrow(obj)
cells_removed_qc <- cells_initial - cells_after_qc
pct_removed_qc <- 100 * cells_removed_qc / cells_initial
```

The threshold-based QC stage retained `r fmt_int(cells_after_qc)` cells after removing low-feature, high-mitochondrial, and obvious outlier barcodes.

## Doublet Detection

```{r doublet-detection}
sce <- as.SingleCellExperiment(obj)
sce$sample <- obj$sample

sce <- scDblFinder::scDblFinder(
  sce,
  samples = sce$sample,
  verbose = FALSE
)

obj$doublet_score <- sce$scDblFinder.score
obj$doublet_status <- sce$scDblFinder.class

doublet_calls <- obj@meta.data %>%
  tibble::rownames_to_column("Cell") %>%
  dplyr::select(Cell, sample, group, doublet_status, doublet_score)

doublet_summary <- doublet_calls %>%
  dplyr::count(doublet_status, name = "cells") %>%
  dplyr::mutate(prop = cells / sum(cells))

doublet_by_sample <- doublet_calls %>%
  dplyr::count(sample, group, doublet_status, name = "cells") %>%
  dplyr::group_by(sample, group) %>%
  tidyr::complete(doublet_status = c("singlet", "doublet"), fill = list(cells = 0L)) %>%
  dplyr::ungroup() %>%
  tidyr::pivot_wider(
    names_from = doublet_status,
    values_from = cells,
    values_fill = 0
  ) %>%
  dplyr::mutate(
    total_cells = singlet + doublet,
    doublet_rate = doublet / total_cells
  ) %>%
  dplyr::arrange(group, sample)

obj <- subset(obj, subset = doublet_status == "singlet")

cells_filtered <- ncol(obj)
cells_removed_doublets <- cells_after_qc - cells_filtered
pct_removed_doublets <- 100 * cells_removed_doublets / cells_after_qc
```

Likely doublets were removed before clustering. This retained `r fmt_int(cells_filtered)` singlet cells for atlas rebuilding.

## Gene Filtering and Normalisation

```{r gene-filtering}
genes_before <- nrow(obj)

counts_mat <- GetAssayData(obj, layer = "counts")
keep_genes <- Matrix::rowSums(counts_mat > 0) >= 10
obj <- subset(obj, features = rownames(obj)[keep_genes])

genes_filtered <- nrow(obj)
genes_removed <- genes_before - genes_filtered
```

```{r normalization}
obj <- NormalizeData(
  obj,
  normalization.method = "LogNormalize",
  scale.factor = 10000
)

obj <- FindVariableFeatures(
  obj,
  selection.method = "vst",
  nfeatures = 2000
)

variable_plot <- VariableFeaturePlot(obj)
save_plot(variable_plot, "atlas_variable_features.png", width = 8, height = 5)
variable_plot
```

## Global Atlas Rebuild

### Fresh Scaling and PCA

```{r scaling-pca}
obj <- ScaleData(
  obj,
  features = VariableFeatures(obj),
  verbose = FALSE
)

obj <- RunPCA(
  obj,
  features = VariableFeatures(obj),
  npcs = 30,
  verbose = FALSE
)

candidate_grid <- tibble::tribble(
  ~pc, ~res,
  15, 0.4,
  20, 0.2,
  20, 0.4,
  20, 0.6,
  25, 0.4
) %>%
  dplyr::mutate(solution_id = paste0("pc", pc, "_res", res))

anchor_solution_id <- "pc20_res0.4"
final_solution_id <- anchor_solution_id  # Working anchor, not an evidence-selected optimum.

pca_plot <- DimPlot(
  obj,
  reduction = "pca",
  dims = c(1, 2),
  group.by = "group",
  raster = TRUE
) +
  ggtitle("PCA of PBMC scRNA-seq Data") +
  theme_classic()

save_plot(pca_plot, "atlas_pca_grouped.png", width = 8, height = 6)
pca_plot

elbow_plot <- ElbowPlot(obj, ndims = 30) +
  scale_x_continuous(breaks = seq(1, 30, by = 2)) +
  ggtitle("Principal Component Selection (Retuning Screen)") +
  xlab("Principal Component") +
  ylab("Standard Deviation") +
  theme_classic()

save_plot(elbow_plot, "atlas_elbow_plot.png", width = 7, height = 5)
elbow_plot
```

```{r pca-by-sample}
pca_sample_plot <- DimPlot(
  obj,
  reduction = "pca",
  dims = c(1, 2),
  group.by = "sample",
  raster = TRUE
) +
  ggtitle("PCA coloured by sample") +
  theme_classic()

save_plot(pca_sample_plot, "atlas_pca_sample.png", width = 8, height = 6)
pca_sample_plot
```

### SingleR Support

```{r singler-support}
ref <- celldex::HumanPrimaryCellAtlasData()

pred <- SingleR(
  test = GetAssayData(obj, layer = "data"),
  ref = ref,
  labels = ref$label.main
)

obj$SingleR_labels <- pred$labels
obj$SingleR_main <- gsub("_", " ", obj$SingleR_labels)
```

### Candidate Solution Helpers

```{r helper-functions}
pbmc_marker_sets <- list(
  t_cells = c("CD3D", "IL7R", "LTB", "MAL", "LEF1", "TRAT1"),
  cytotoxic = c("NKG7", "GNLY", "KLRD1", "TRAC", "CD8A", "GZMK", "PRF1", "KLRF1", "TRDC"),
  b_cells = c("MS4A1", "CD79A", "CD74", "HLA-DRA", "TCL1A", "IGHM", "IGHD", "FCRLA"),
  monocytes = c("LYZ", "LST1", "S100A8", "S100A12", "FCGR3A", "CTSD", "SAT1", "C1QA", "LGALS2", "THBD"),
  cdc = c("FCER1A", "CD1C", "CLEC10A", "CST3"),
  pdc = c("CLEC4C", "LILRA4", "TCF4", "JCHAIN", "GZMB"),
  platelets = c("PPBP", "PF4", "GP9", "CLEC1B", "ESAM", "SDPR"),
  erythroid = c("HBA1", "HBA2", "HBB", "GYPA", "ALAS2", "AHSP", "CA1", "GYPB"),
  ifn_high = c("IFI27", "ISG15", "IFITM3", "IFIT1", "IFIT3", "MX1", "OAS1")
)

count_marker_hits <- function(marker_genes, marker_set) {
  sum(unique(unlist(marker_genes)) %in% marker_set)
}

classify_cluster <- function(marker_genes, max_sample_frac, n_cells) {
  platelet_hits <- count_marker_hits(marker_genes, pbmc_marker_sets$platelets)
  erythroid_hits <- count_marker_hits(marker_genes, pbmc_marker_sets$erythroid)
  ifn_hits <- count_marker_hits(marker_genes, pbmc_marker_sets$ifn_high)
  pdc_hits <- count_marker_hits(marker_genes, pbmc_marker_sets$pdc)
  cdc_hits <- count_marker_hits(marker_genes, pbmc_marker_sets$cdc)
  b_hits <- count_marker_hits(marker_genes, pbmc_marker_sets$b_cells)
  monocyte_hits <- count_marker_hits(marker_genes, pbmc_marker_sets$monocytes)
  cytotoxic_hits <- count_marker_hits(marker_genes, pbmc_marker_sets$cytotoxic)
  tcell_hits <- count_marker_hits(marker_genes, pbmc_marker_sets$t_cells)
  nonclassical_hits <- sum(unique(marker_genes) %in% c("FCGR3A", "CTSD", "SAT1", "C1QA", "LST1", "CSF1R"))
  activated_mono_hits <- sum(unique(marker_genes) %in% c("S100A8", "S100A12", "VCAN", "LGALS2", "THBD", "EREG", "FCN1", "MCEMP1"))

  suggested_label <- dplyr::case_when(
    platelet_hits >= 2 ~ "Platelet-rich contaminant",
    erythroid_hits >= 2 ~ "Erythroid contaminant",
    ifn_hits >= 2 ~ "IFN-high mixed cells",
    pdc_hits >= 2 ~ "Plasmacytoid dendritic cells",
    cdc_hits >= 2 ~ "Conventional dendritic cells",
    b_hits >= 2 ~ "B cells",
    nonclassical_hits >= 2 ~ "Non-classical monocytes",
    activated_mono_hits >= 2 | monocyte_hits >= 2 ~ "Activated monocytes",
    tcell_hits >= 2 & cytotoxic_hits < 2 ~ "IL7R+ T cells",
    cytotoxic_hits >= 2 ~ "Cytotoxic lymphocytes",
    TRUE ~ "Ambiguous cells"
  )

  confidence <- dplyr::case_when(
    suggested_label %in% c("Platelet-rich contaminant", "Erythroid contaminant", "IFN-high mixed cells", "Ambiguous cells") ~ "Low",
    max_sample_frac > 0.80 ~ "Low",
    max_sample_frac > 0.50 ~ "Medium",
    n_cells < 150 ~ "Medium",
    TRUE ~ "High"
  )

  notes <- c()

  if (max_sample_frac > 0.80) {
    notes <- c(notes, "Library-private cluster; interpret cautiously.")
  } else if (max_sample_frac > 0.50) {
    notes <- c(notes, "Library-enriched cluster; requires reproducibility check.")
  }

  if (n_cells < 150) {
    notes <- c(notes, "Small cluster; marker stability may be limited.")
  }

  if (suggested_label == "Platelet-rich contaminant") {
    notes <- c(notes, "Keep visible in the atlas but quarantine from later immune interpretation.")
  }

  if (suggested_label == "Erythroid contaminant") {
    notes <- c(notes, "Do not interpret as proliferating immune cells.")
  }

  if (suggested_label == "IFN-high mixed cells") {
    notes <- c(notes, "Keep only if this state persists across candidate solutions and multiple libraries.")
  }

  if (suggested_label == "Ambiguous cells") {
    notes <- c(notes, "Mixed or weak lineage markers; avoid over-annotation.")
  }

  tibble::tibble(
    suggested_label = suggested_label,
    confidence = confidence,
    notes = paste(notes, collapse = " ")
  )
}

run_candidate_solution <- function(seu, pc, res, solution_id) {
  tmp <- FindNeighbors(seu, dims = 1:pc, verbose = FALSE)
  tmp <- FindClusters(tmp, resolution = res, random.seed = 1234, verbose = FALSE)
  tmp <- RunUMAP(tmp, dims = 1:pc, seed.use = 1234, verbose = FALSE)

  tmp$cluster_screen <- as.character(Idents(tmp))
  Idents(tmp) <- "cluster_screen"

  future::plan(multisession, workers = marker_workers)
  markers_screen <- FindAllMarkers(
    tmp,
    only.pos = TRUE,
    min.pct = 0.25,
    logfc.threshold = 0.25,
    max.cells.per.ident = 400,
    verbose = FALSE
  ) %>%
    dplyr::filter(p_val_adj < 0.05)
  future::plan(sequential)

  top10_screen <- markers_screen %>%
    dplyr::group_by(cluster) %>%
    dplyr::slice_max(avg_log2FC, n = 10, with_ties = FALSE) %>%
    dplyr::ungroup()

  library_contrib <- tmp@meta.data %>%
    as.data.frame() %>%
    tibble::rownames_to_column("Cell") %>%
    tibble::as_tibble() %>%
    dplyr::mutate(cluster = as.character(cluster_screen)) %>%
    dplyr::count(cluster, sample, group, name = "cells") %>%
    dplyr::group_by(cluster) %>%
    dplyr::mutate(cluster_total = sum(cells), frac_within_cluster = cells / cluster_total) %>%
    dplyr::ungroup()

  cluster_summary <- library_contrib %>%
    dplyr::group_by(cluster) %>%
    dplyr::summarise(
      cells = max(cluster_total),
      n_samples = dplyr::n_distinct(sample),
      dominant_sample = sample[which.max(frac_within_cluster)],
      dominant_group = group[which.max(frac_within_cluster)],
      max_sample_frac = max(frac_within_cluster),
      .groups = "drop"
    )

  singleR_support <- tmp@meta.data %>%
    as.data.frame() %>%
    tibble::rownames_to_column("Cell") %>%
    tibble::as_tibble() %>%
    dplyr::mutate(cluster = as.character(cluster_screen)) %>%
    dplyr::count(cluster, SingleR_main, name = "cells") %>%
    dplyr::group_by(cluster) %>%
    dplyr::mutate(frac = cells / sum(cells)) %>%
    dplyr::slice_max(cells, n = 1, with_ties = FALSE) %>%
    dplyr::ungroup() %>%
    dplyr::transmute(cluster, singleR_majority = paste0(SingleR_main, " (", round(100 * frac, 1), "%)"))

  cluster_review <- top10_screen %>%
    dplyr::group_by(cluster) %>%
    dplyr::summarise(
      top_markers = paste(gene[seq_len(min(5, dplyr::n()))], collapse = ", "),
      marker_vector = list(gene),
      .groups = "drop"
    ) %>%
    dplyr::right_join(cluster_summary, by = "cluster") %>%
    dplyr::left_join(singleR_support, by = "cluster") %>%
    dplyr::mutate(
      classification = purrr::pmap(list(marker_vector, max_sample_frac, cells), classify_cluster)
    ) %>%
    tidyr::unnest_wider(classification) %>%
    dplyr::mutate(solution_id = solution_id, pc = pc, res = res)

  solution_summary <- cluster_review %>%
    dplyr::summarise(
      solution_id = solution_id[1],
      pc = pc[1],
      res = res[1],
      n_clusters = dplyr::n(),
      library_enriched_clusters = sum(max_sample_frac > 0.50),
      library_private_clusters = sum(max_sample_frac > 0.80),
      quarantined_or_low_confidence = sum(confidence == "Low"),
      .groups = "drop"
    )

  umap_plot <- DimPlot(
    tmp,
    reduction = "umap",
    label = TRUE,
    repel = TRUE,
    raster = TRUE
  ) +
    ggtitle(paste0("Candidate solution: ", solution_id)) +
    theme_classic()

  list(
    summary = solution_summary,
    review = cluster_review,
    top10 = top10_screen,
    umap_plot = umap_plot
  )
}
```

### Candidate Clustering Solutions

```{r candidate-solutions}
candidate_results <- purrr::pmap(
  candidate_grid,
  function(pc, res, solution_id) run_candidate_solution(obj, pc, res, solution_id)
)

names(candidate_results) <- candidate_grid$solution_id

candidate_solution_overview <- dplyr::bind_rows(
  purrr::map(candidate_results, "summary")
) %>%
  dplyr::arrange(pc, res)

candidate_cluster_review <- dplyr::bind_rows(
  purrr::map(candidate_results, "review")
) %>%
  dplyr::arrange(solution_id, cluster)

candidate_top10_markers <- dplyr::bind_rows(
  purrr::imap(candidate_results, ~ dplyr::mutate(.x$top10, solution_id = .y))
)

candidate_umap_panel <- patchwork::wrap_plots(
  purrr::map(candidate_results, "umap_plot"),
  ncol = 2
)

save_plot(candidate_umap_panel, "atlas_candidate_umap_solutions.png", width = 12, height = 10)

knitr::kable(
  candidate_solution_overview,
  digits = 3,
  caption = "Candidate clustering solutions screened during atlas retuning."
)
```

```{r candidate-umap-panel}
candidate_umap_panel
```

```{r anchor-solution-review}
anchor_cluster_review <- candidate_cluster_review %>%
  dplyr::filter(solution_id == anchor_solution_id) %>%
  dplyr::select(
    cluster,
    cells,
    top_markers,
    dominant_sample,
    dominant_group,
    max_sample_frac,
    singleR_majority,
    suggested_label,
    confidence,
    notes
  )

knitr::kable(
  anchor_cluster_review,
  digits = 3,
  caption = "Cluster review table for the anchor solution."
)
```

### Final Atlas Solution

```{r final-solution}
stopifnot(final_solution_id %in% candidate_grid$solution_id)

final_pc <- candidate_grid$pc[candidate_grid$solution_id == final_solution_id][1]
final_res <- candidate_grid$res[candidate_grid$solution_id == final_solution_id][1]

obj <- FindNeighbors(obj, dims = 1:final_pc, verbose = FALSE)
obj <- FindClusters(obj, resolution = final_res, random.seed = 1234, verbose = FALSE)
obj <- RunUMAP(obj, dims = 1:final_pc, seed.use = 1234, verbose = FALSE)

obj$cluster_final <- as.character(Idents(obj))
Idents(obj) <- "cluster_final"
```

The working anchor clustering was frozen as `cluster_final` before annotation. Numeric clusters were treated as the primary unit of interpretation throughout atlas review.

### Final Markers and Library Contribution

```{r final-markers}
future::plan(multisession, workers = marker_workers)

markers <- FindAllMarkers(
  obj,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25,
  verbose = FALSE
) %>%
  dplyr::filter(p_val_adj < 0.05)

future::plan(sequential)

top10_markers <- markers %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_max(avg_log2FC, n = 10, with_ties = FALSE) %>%
  dplyr::ungroup()

top5_markers <- markers %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_max(avg_log2FC, n = 5, with_ties = FALSE) %>%
  dplyr::ungroup()

sample_contribution <- obj@meta.data %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Cell") %>%
  tibble::as_tibble() %>%
  dplyr::mutate(cluster = as.character(cluster_final)) %>%
  dplyr::count(cluster, sample, group, name = "cells") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(cluster_total = sum(cells), frac_within_cluster = cells / cluster_total) %>%
  dplyr::ungroup()

cluster_library_summary <- sample_contribution %>%
  dplyr::group_by(cluster) %>%
  dplyr::summarise(
    cells = max(cluster_total),
    samples_detected = dplyr::n_distinct(sample),
    dominant_sample = sample[which.max(frac_within_cluster)],
    dominant_group = group[which.max(frac_within_cluster)],
    max_sample_frac = max(frac_within_cluster),
    .groups = "drop"
  ) %>%
  dplyr::arrange(as.numeric(cluster))

knitr::kable(
  cluster_library_summary,
  digits = 3,
  caption = "Library contribution summary for each final numeric cluster."
)
```

### Conservative Annotation Table

```{r annotation-table}
singleR_support <- obj@meta.data %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Cell") %>%
  tibble::as_tibble() %>%
  dplyr::mutate(cluster = as.character(cluster_final)) %>%
  dplyr::count(cluster, SingleR_main, name = "cells") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(frac = cells / sum(cells)) %>%
  dplyr::slice_max(cells, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::transmute(cluster, singleR_majority = paste0(SingleR_main, " (", round(100 * frac, 1), "%)"))

annotation_table <- top10_markers %>%
  dplyr::group_by(cluster) %>%
  dplyr::summarise(
    top_markers = paste(gene[seq_len(min(5, dplyr::n()))], collapse = ", "),
    marker_vector = list(gene),
    .groups = "drop"
  ) %>%
  dplyr::right_join(cluster_library_summary, by = "cluster") %>%
  dplyr::left_join(singleR_support, by = "cluster") %>%
  dplyr::mutate(
    classification = purrr::pmap(list(marker_vector, max_sample_frac, cells), classify_cluster)
  ) %>%
  tidyr::unnest_wider(classification) %>%
  dplyr::mutate(
    notes = stringr::str_trim(
      paste0(
        notes,
        ifelse(notes == "", "", " "),
        "SingleR support: ",
        singleR_majority
      )
    )
  ) %>%
  dplyr::transmute(
    cluster,
    final_label = suggested_label,
    top_markers,
    confidence,
    notes
  ) %>%
  dplyr::arrange(as.numeric(cluster))

annotation_overrides <- tibble::tibble(
  cluster = character(),
  final_label = character(),
  confidence = character(),
  notes = character()
)
# Add reviewed cluster-level overrides here after inspecting markers and library contribution.

if (nrow(annotation_overrides) > 0) {
  annotation_table <- annotation_table %>%
    dplyr::left_join(annotation_overrides, by = "cluster", suffix = c("", ".override")) %>%
    dplyr::mutate(
      final_label = dplyr::coalesce(final_label.override, final_label),
      confidence = dplyr::coalesce(confidence.override, confidence),
      notes = dplyr::coalesce(notes.override, notes)
    ) %>%
    dplyr::select(cluster, final_label, top_markers, confidence, notes)
}

label_lookup <- setNames(annotation_table$final_label, annotation_table$cluster)
obj$cell_type_final <- dplyr::recode(as.character(obj$cluster_final), !!!label_lookup)

knitr::kable(
  annotation_table,
  caption = "Provisional annotation table for the final global PBMC atlas."
)
```

```{r suspicious-clusters}
suspicious_clusters <- annotation_table %>%
  dplyr::filter(
    grepl("contaminant|IFN-high|Ambiguous", final_label) |
      grepl("Library-private|Library-enriched", notes)
  )

knitr::kable(
  suspicious_clusters,
  caption = "Clusters requiring quarantine or especially cautious interpretation."
)
```

### Final Atlas Outputs

```{r umap-numeric}
Idents(obj) <- "cluster_final"

umap_clusters <- DimPlot(
  obj,
  reduction = "umap",
  label = TRUE,
  repel = TRUE,
  raster = TRUE
) +
  ggtitle("Final UMAP coloured by numeric cluster") +
  theme_classic()

save_plot(umap_clusters, "atlas_umap_clusters_final.png", width = 8, height = 6)
umap_clusters
```

```{r umap-celltype}
Idents(obj) <- "cell_type_final"

annotated_umap <- DimPlot(
  obj,
  reduction = "umap",
  label = TRUE,
  repel = TRUE,
  raster = TRUE
) +
  ggtitle("Final UMAP coloured by conservative atlas labels") +
  theme_classic()

save_plot(annotated_umap, "atlas_umap_celltypes_final.png", width = 9, height = 6)
annotated_umap
```

```{r umap-sample}
umap_sample <- DimPlot(
  obj,
  reduction = "umap",
  group.by = "sample",
  raster = TRUE
) +
  ggtitle("Final UMAP coloured by sample") +
  theme_classic()

save_plot(umap_sample, "atlas_umap_sample_final.png", width = 9, height = 6)
umap_sample
```

```{r canonical-dotplot}
DefaultAssay(obj) <- "RNA"
Idents(obj) <- "cluster_final"

canonical_dotplot <- DotPlot(
  obj,
  group.by = "cluster_final",
  features = c(
    "CD3D", "IL7R", "LTB", "MAL",
    "NKG7", "GNLY", "KLRD1", "TRAC", "CD8A", "GZMK", "PRF1",
    "MS4A1", "CD79A", "CD74", "HLA-DRA", "TCL1A", "IGHM", "IGHD",
    "LYZ", "LST1", "S100A8", "S100A12", "FCGR3A", "CTSD", "SAT1", "C1QA",
    "FCER1A", "CD1C", "CLEC10A", "CST3",
    "CLEC4C", "LILRA4", "TCF4", "JCHAIN", "GZMB",
    "PPBP", "PF4", "GP9", "CLEC1B",
    "HBA1", "HBA2", "HBB", "GYPA", "ALAS2"
  )
) +
  RotatedAxis()

save_plot(canonical_dotplot, "atlas_canonical_marker_dotplot.png", width = 14, height = 8)
canonical_dotplot
```

```{r marker-heatmap}
top_heatmap_genes <- top10_markers %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_max(avg_log2FC, n = 5, with_ties = FALSE) %>%
  dplyr::pull(gene) %>%
  unique()

cells_use <- unlist(lapply(
  split(colnames(obj), obj$cluster_final),
  function(x) sample(x, min(50, length(x)))
))

obj <- ScaleData(obj, features = top_heatmap_genes, verbose = FALSE)

marker_heatmap <- DoHeatmap(
  obj,
  features = top_heatmap_genes,
  cells = cells_use,
  group.by = "cluster_final"
) + NoLegend()

save_plot(marker_heatmap, "atlas_marker_heatmap.png", width = 11, height = 8)
marker_heatmap
```

```{r sample-contribution-plot}
sample_contribution_plot <- sample_contribution %>%
  dplyr::mutate(
    sample = factor(sample, levels = unique(sample[order(group, sample)])),
    cluster = factor(cluster, levels = sort(unique(cluster)))
  ) %>%
  ggplot(aes(x = cluster, y = sample, fill = frac_within_cluster)) +
  geom_tile(color = "white") +
  facet_grid(group ~ ., scales = "free_y", space = "free_y") +
  scale_fill_gradient(low = "grey95", high = "firebrick") +
  xlab("Numeric cluster") +
  ylab("Sample") +
  ggtitle("Sample contribution to each final numeric cluster") +
  theme_classic() +
  theme(
    panel.spacing = grid::unit(0.3, "lines"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

save_plot(sample_contribution_plot, "atlas_sample_contribution_by_cluster.png", width = 10, height = 8)
sample_contribution_plot
```

```{r cluster-sizes}
cluster_sizes <- annotation_table %>%
  dplyr::right_join(cluster_library_summary, by = "cluster") %>%
  dplyr::transmute(
    Cluster = cluster,
    Final_label = final_label,
    Cells = cells,
    Proportion = sprintf("%.1f%%", 100 * cells / sum(cells))
  ) %>%
  dplyr::arrange(desc(Cells))

n_clusters <- nrow(cluster_sizes)

knitr::kable(
  cluster_sizes,
  caption = "Cluster sizes for the final global PBMC atlas."
)
```

## Provisional atlas summary

The atlas rebuild retained `r fmt_int(cells_filtered)` singlet cells and `r fmt_int(genes_filtered)` genes for downstream global mapping. Candidate clustering solutions were screened across 15 to 25 PCs and resolutions from 0.2 to 0.6, with `r final_pc` PCs and resolution `r final_res` used as a provisional anchor. The code does not automatically establish an optimal or stable solution.

This report intentionally stops at the global atlas stage. It does not perform composition testing, differential expression, pathway enrichment, subclustering, trajectory inference, RNA velocity, or cell-cell communication. Those analyses should wait until the global cluster definitions and conservative annotation table are judged stable and reproducible.

## Exported Outputs

```{r exports}
write.csv(doublet_calls, file.path(exports_dir, "doublet_calls.csv"), row.names = FALSE)
write.csv(doublet_summary, file.path(exports_dir, "doublet_summary.csv"), row.names = FALSE)
write.csv(doublet_by_sample, file.path(exports_dir, "doublet_by_sample.csv"), row.names = FALSE)

write.csv(candidate_solution_overview, file.path(exports_dir, "candidate_solution_overview.csv"), row.names = FALSE)
write.csv(csv_ready(candidate_cluster_review), file.path(exports_dir, "candidate_cluster_review.csv"), row.names = FALSE)
write.csv(candidate_top10_markers, file.path(exports_dir, "candidate_top10_markers.csv"), row.names = FALSE)

write.csv(markers, file.path(exports_dir, "all_markers.csv"), row.names = FALSE)
write.csv(top10_markers, file.path(exports_dir, "top10_markers_by_cluster.csv"), row.names = FALSE)
write.csv(top5_markers, file.path(exports_dir, "top5_markers_by_cluster.csv"), row.names = FALSE)
write.csv(cluster_library_summary, file.path(exports_dir, "cluster_library_summary.csv"), row.names = FALSE)
write.csv(sample_contribution, file.path(exports_dir, "sample_contribution_by_cluster.csv"), row.names = FALSE)
write.csv(cluster_sizes, file.path(exports_dir, "cluster_sizes.csv"), row.names = FALSE)
write.csv(annotation_table, file.path(exports_dir, "cluster_annotation_table.csv"), row.names = FALSE)
write.csv(suspicious_clusters, file.path(exports_dir, "suspicious_clusters.csv"), row.names = FALSE)
write.csv(
  data.frame(Cell = colnames(obj), SingleR_label = obj$SingleR_main),
  file.path(exports_dir, "singleR_labels.csv"),
  row.names = FALSE
)
write.csv(
  obj@meta.data,
  file.path(exports_dir, "cell_metadata_with_atlas_labels.csv"),
  row.names = TRUE
)

saveRDS(obj, file.path(exports_dir, "pbmc_global_atlas_seurat.rds"))
```

## Session Info

```{r session-info}
writeLines(capture.output(sessionInfo()), file.path(exports_dir, "session_info.txt"))
saveRDS(obj, file.path(exports_dir, "analysis_seurat.rds"))
sessionInfo()
```
